---
title: "metagene: RNA-Seq version (experimental extension)"
output:
    BiocStyle::html_document:
        toc: true
vignette: >
    %\VignetteIndexEntry{Introduction to metagene's RNA-Seq analysis}
    %\VignetteEngine{knitr::rmarkdown}
    \usepackage[utf8]{inputenc}
---

```{r style, echo = FALSE, results = 'asis', message = FALSE}
BiocStyle::markdown()
library(knitr)
```

**Package**: `r Biocpkg("metagene")`<br />
**Modified**: 17 october, 2017<br />
**Compiled**: `r date()`<br />
**License**: `r packageDescription("metagene")[["License"]]`<br />

# Introduction

This experimental functionality of metagene extends the use metagene to RNA-Seq
data analysis. Gene-level quantification of the RNA-seq experiment is performed
directly from coverage profiles, and differential expression can be assessed
using the similaRpeak package.
Experimental designs can account for multiple replicates divided into several
groups of samples. Similarly to the analysis of ChIP-Seq data by metagene, 
bootstrap procedures are used to obtain a better estimate of the mean enrichment 
and of the confidence interval for every group of samples.

This vignette introduces all of the main features of the metagene RNA-seq 
extension.

# Loading the metagene package

```{r libraryLoad, message = FALSE}
library(metagene)
```

# Inputs

## Alignment files (BAM files)

There is no hard limit to the number of BAM files that can be included in an
analysis, though memory usage might become an issue when large numbers of BAM 
files are analyzed at once. All provided BAM files must be indexed: the BAM
`file.bam` should have an accompanying `file.bam.bai` or `file.bai` index file
in the same directory.

The paths (relative or absolute) to the BAM files must be in a vector:
```{r bamFiles}
bam_files <- 
c(system.file("extdata/cyto4.bam", package="metagene"),
  system.file("extdata/cyto3.bam", package="metagene"),
  system.file("extdata/nuc4.bam", package="metagene"),
  system.file("extdata/nuc3.bam", package="metagene"))
```
BAM file from both single-end of paired-end RNA-seq experiments might be used
(see below for details)

## Genomic regions

### BED files

To compare custom regions of interest, it is possible to use a list of one or
more BED files. Here, two genes are used. The name of the files (without the 
extension) will be used to name each gene. 
**Each gene of interest must be in its own BED file.** Ranges within 
the BED files will be interpreted as exons within that gene. For example, 
see the 'DPM1.bed' BED file in 'metagene/inst/extdata'.

```{r regionsArgument}
regions <- 
c(system.file("extdata/DPM1.bed", package="metagene"),
system.file("extdata/NDUFAB1.bed", package="metagene"))
```
### GRanges or GRangesList objects - Regions

As an alternative to a list of BED files, An object of type `GRangesList`
can be used to specify genes of interest. Each element of the `GRangesList` must 
be a gene of interest, its component `GRanges` describing its exons.

## Design groups

An experimental design describes the set of BAM files to be used
during the metagene analysis, and their relationships to one another.
An experimental design's first column should provide the paths to the BAM files
to be analyzed. Subsequent columns should describe design groups using the
following values:
    + 0: ignore file
    + 1: input
    + 2: control

As an example, a design could be created directly from R code:

```{r DesignCraft}
Samples <- c("cyto4.bam",
            "cyto3.bam",
            "nuc4.bam",
            "nuc3.bam")
cyto <- c(1,1,0,0)
nucleo <- c(0,0,1,1)
mydesign <- cbind(Samples, cyto, nucleo)
mydesign <- data.frame(mydesign)
#to make cyto and nucleo colums numeric variables
mydesign$cyto <- cyto
mydesign$nucleo <- nucleo
```

# Analysis steps

A typical metagene analysis will consist of the following steps:

* Extraction of the read counts of every BAM files in selected regions.
* Conversion into coverages.
* Normalization of the coverage values.
* Table production.
* Data frame production.
* Generation of the metagene plot.
* Profile comparison (differential expression)

## Creation of a metagene object for an RNA-seq analysis

Two new arguments are introduced to the metagene constructor for RNA-Seq support:
paired_end and assay. Data used in the following example were obtained 
from the ENCODE experiments ENCSR000CPJ and ENCSR000CPK, using paired-end 
polyA mRNA RNA-seq.

```{r new metagene_rnaseq object, warnings = FALSE}
bam_files <- 
c(system.file("extdata/nuc4.bam", package="metagene"))
regions <- 
c(system.file("extdata/DPM1.bed", package="metagene"))
# Initialization
mg <- metagene$new(regions = regions, 
                    bam_files = bam_files, 
                    assay = 'rnaseq', 
                    paired_end = TRUE)
mg$produce_table(normalization = 'RPM')
mg$produce_data_frame()
mg$plot(title = 'DPM1')
```
The plot shows the coverage in Reads-per-million(RPM) of the gene DPM1 
for the bam file nuc4.bam. The x-axis is orientated from 3' to 5' (from 0 from 1250)
because of the strand orientation of the gene in the input BED file.

However, you can control normalization, flip and other arguments provided in 
produce_table and produce_data_frame methods. Let's see that together.
The dashed lines represent exon frontiers.

NOTE: in paired_end, a warning message could appear to warn you about how many 
alignments with ambiguous pairing were dumped.

It is not mandatory to explicitly call each step of the 
metagene analysis. For instance, in the previous example, the `plot` function 
could have been called directly, and would have called all other intermediary 
steps automatically using default values. The next section will describe those
steps in more details.



## Complete analysis

In order to fully control every step of a metagene analysis in RNA-Seq version, 
it is important to understand how a complete analysis is performed. If we are 
satisfied with the default values, it is not mandatory to explicitly call every 
step (as was shown in the previous section).

### Initialization

During this step, the coverages for every regions specified are extracted from
every BAM files. Let's start with a bigger dataset as introduced in inputs 
section. Here, we will use four bam files and three genes.

```{r initialization, warnings = FALSE}
bam_files <- 
c(system.file("extdata/cyto4.bam", package="metagene"),
  system.file("extdata/cyto3.bam", package="metagene"),
  system.file("extdata/nuc4.bam", package="metagene"),
  system.file("extdata/nuc3.bam", package="metagene"))
regions <- 
c(system.file("extdata/DPM1.bed", package="metagene"),
system.file("extdata/NDUFAB1.bed", package="metagene"))
mg <- metagene$new(regions = regions, 
                    bam_files = bam_files, 
                    assay = 'rnaseq', 
                    paired_end = TRUE)
```

### Producing the table

No new parameters was added to the produce_table method. However, new colums to 
the table were added but is completely transparent for user. Colums names are now :
`r colnames(mg$get_table())`. Regions was kept and not replaced by gene but
stands for gene names. Data were not binned as for ChIP-Seq analysis and bin 
column was replaced by nuc (for nucleotide) column.

```{r showProduceTable, warnings = FALSE}
mg$produce_table(flip_regions = TRUE, normalization = 'RPM')
tab <- mg$get_table()
```
Here the flip argument is `TRUE`. Thus, all genes will be oriented from 5' to 3'
on the metagene plot. The 'Reads Per Millions' (RPM) normalization was done.

### Producing the `data.frame`

The metagene plot are produced using the `ggplot2` package, which require a
`data.frame` as input. During this step, the values of the ribbon are
calculated. Metagene RNA-Seq version uses "bootstrap" to obtain a better 
estimation of the mean coverage for every positions in each group samples.

```{r produceDataFrame}
mg$produce_data_frame()
```

An additionnal option was implemented into the produce_data_frame method. The 
possibility to remove values which are under a certain threshold. This option 
is used in order to remove, for instance, exon or part of exon for which there 
is no data and thereby obtain a profile without gaps. There are the new 
arguments to do that : (1) 'avoid_gaps' a logical argument to allow this 
removal or not, (2) 'bam_name' a character argument that give the name of the 
bam file to take as reference for values to remove also for other samples 
(default : the first sample in table) (3) 'gaps_threshold' a numeric threshold 
under which value will be removed from data_frame (default = 0).

```{r produceDataFrame2, warnings = FALSE}
bam_files <- 
c(system.file("extdata/cyto4.bam", package="metagene"))
region <- 
c(system.file("extdata/NDUFAB1.bed", package="metagene"))
mg <- metagene$new(regions = region, bam_files = bam_files, assay = 'rnaseq')
mg$produce_table(flip_regions = TRUE, normalization = 'RPM')
mg$plot(title = 'with all normalized values')
mg$produce_data_frame(avoid_gaps = TRUE,
                        bam_name = "cyto4",
                        gaps_threshold = 30)
mg$plot(title = 'without normalized values under 30')
```

### Plotting

During this step, metagene will use the `data.frame` to plot the calculated
values using `ggplot2`. A subset of the gene and design can be selected by using
 the `region_names` and `design_names` arguments. The `region_names` correspond 
to the names of the genes used during the initialization. The `design_name` 
will vary depending if a design was added. If no design was added, this param 
correspond to the BAM name or BAM filenames. Otherwise, we have to use the 
names of the columns from the design.

```{r showPlot, warnings = FALSE}
bam_files <- 
c(system.file("extdata/cyto4.bam", package="metagene"),
  system.file("extdata/cyto3.bam", package="metagene"),
  system.file("extdata/nuc4.bam", package="metagene"),
  system.file("extdata/nuc3.bam", package="metagene"))
regions <- 
c(system.file("extdata/DPM1.bed", package="metagene"),
system.file("extdata/NDUFAB1.bed", package="metagene"))
mg <- metagene$new(regions = regions, 
                    bam_files = bam_files, 
                    assay = 'rnaseq', 
                    paired_end = TRUE)
mg$produce_table(normalization = 'RPM')
mg$produce_data_frame(avoid_gaps = TRUE,
                        bam_name = "cyto4",
                        gaps_threshold = 30)
mg$plot(region_names = "DPM1", title = "Demo plot subset RNA-Seq")
```

# Managing large datasets

While `metagene` try to reduce it's memory usage, it's possible to run into
memory limits when working with multiple large datasets (especially when there
is a lot of genes).

One way to avoid this is to analyse each dataset seperately and then merge just
before producing the metagene plot:

```{r memory, collapse=TRUE, warnings = FALSE, eval=FALSE}
regions <- 
c(system.file("extdata/DPM1.bed", package="metagene"),
system.file("extdata/NDUFAB1.bed", package="metagene"))
bam_files <- 
c(system.file("extdata/cyto4.bam", package="metagene"),
  system.file("extdata/cyto3.bam", package="metagene"),
  system.file("extdata/nuc4.bam", package="metagene"),
  system.file("extdata/nuc3.bam", package="metagene"))
mg1 <- metagene$new(bam_files = bam_files[1:2], regions = regions, 
							assay = 'rnaseq', paired_end = TRUE)
mg1$produce_data_frame()
mg2 <- metagene$new(bam_files = bam_files[3:4], regions = regions, 
							assay = 'rnaseq', paired_end = TRUE)
mg2$produce_data_frame()
```

Then you can extract the `data.frame`s and combine them with `rbind`:

```{r extractDF, eval=FALSE}
df1 <- mg1$get_data_frame()
df2 <- mg2$get_data_frame()
df <- rbind(df1, df2)
```

Finally, you can use the `plot_metagene` function to produce the metagene plot:

```{r out.width="px", out.height="400px", eval=FALSE}
p <- plot_metagene(df)
p + ggplot2::ggtitle("Managing large datasets")

```
Here three genes are drawn on the same plot. Strong solid lines stand for gene 
separations and dashed lines for exon separations.

# Comparing profiles with permutations

It is possible to compare two metagene profiles using the `permutation_test`
function provided with the `metagene` package. Please note that the permutation
tests functionality is still in development and is expected to change in future
releases.

The first step is to decide which profiles we want to compare and extract the 
corresponding tables :

It is possible to compare two metagene profiles using the `permutation_test`
function provided with the `metagene` package. Please note that the permutation
tests functionality is still in development and is expected to change in future
releases.

As before in produce_data_frame, you get the possibility to remove values under 
a certain threshold. In order to be able to statistically compare these profiles
 underestimated by the values under the threshold, a compagnon function called
'avoid_gaps_update' was created to carry out the same treatment underwent at 
the data_frame level.

The first step is to decide which profiles we want to compare and extract the 
corresponding tables :

```{r extract_subtables, warnings = FALSE}
mg <- metagene$new(bam_files = bam_files, regions = regions, 
							assay = 'rnaseq', paired_end = TRUE)
mg$produce_table(design = mydesign, normalization = 'RPM')
tab <- mg$get_table()
tab0 <- metagene:::avoid_gaps_update(tab, 
       bam_name = 'cyto4', gaps_threshold = 10)
tab0 <- tab0[which(tab0$region == "NDUFAB1"),]
tab1 <- tab0[which(tab0$design == "cyto"),]
tab2 <- tab0[which(tab0$design == "nucleo"),]
```

Then we defined to function to use to compare the two profiles. For this, a
companion package of `metagene` named `r Biocpkg("similaRpeak")` provides
multiple metrics.

For this example, we will prepare a function to calculate the
RATIO_INTERSECT between two profiles:

```{r similaRpeak}
library(similaRpeak)
perm_fun <- function(profile1, profile2) {
    sim <- similarity(profile1, profile2)
    sim[["metrics"]][["RATIO_INTERSECT"]]
}
```

We then compare our two profiles using this metric:

```{r calculateRNI}
ratio_intersect <- 
  perm_fun(tab1[, .(moy=mean(value)), by=nuc]$moy, 
           tab2[, .(moy=mean(value)), by=nuc]$moy)
ratio_intersect
```

To check if this value is significant, we can permute the two tables that
were used to produce the profile and calculate their
RATIO_INTERSECT:

```{r permTest}
permutation_results <- permutation_test(tab1, tab2, sample_size = 2,
                                        sample_count = 1000, FUN = perm_fun)
```
NB : sample_size here is very low to get an accurate permutation test. More 
replicates by design are needed to improve the quality of the result.

Finally, we check how often the calculated value is greater than the results of
the permutations:

```{r perm_pval}
sum(ratio_intersect >= permutation_results) / 
                                length(permutation_results)
mg$plot(region_names = 'NDUFAB1', title='NDUFAB1')
```

